---
title: "Quick start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Working with age group, period, and cohort labels is often awkward and error-prone. The `agetime` package makes it easier. This vignette introduces some of the basic functions. The vignette focuses on age group labels, but equivalent functions also exist for period and cohort labels.

## Infer intervals from labels

Given a vector of age group labels, extract lower limits, upper limits, midpoints, and widths:

```{r extract}
library(agetime)

x <- c("0", "1-4", "5-9", "10+")
age_lower(x)
age_upper(x)
age_mid(x)
age_width(x)
```

Use the lower limit to filter rows:

```{r filter}
library(dplyr, warn.conflicts = FALSE)

df <- tibble(
  age   = c("0-14", "15-64", "65+"),
  count = c(100, 200, 50)
)

df |>
  filter(age_lower(age) >= 15)
```

## Find special labels

`agetime` knows about totals and open intervals.

```{r filter-total-open}
df <- tibble(
  age   = c("15-64", "0-14", "65+", "All"),
  count = c(5, 2, 6, 13)
)

df |>
  filter(!age_is_total(age))

df |>
  filter(age_is_open_right(age))
```

## Standardize messy labels

`age_standard()` converts labels to a standard format:

```{r standard}
x <- c("5to9", "10--14", "100plus")
age_standard(x)
```

## Create a regular series of labels

The `age_labels` functions generate labels from scratch.

```{r labels}
age_labels_five(lower_first = 0, lower_last = 20)
age_labels_life(lower_last = 60) ## life table age groups
```

## Manipulate levels

Use the `age_coarsen` functions to convert to wider age groups.

```{r coarsen}
x <- c("0-4", "5-9", "10-14", "50-54")
age_coarsen(x, breaks = c(0, 15, 55))
age_coarsen_five(x)
```

Use the `age_fill` functions to add intermediate levels:
```{r fill}
x |>
  age_coarsen_five() |>
  age_fill_five()
```

## Check and assert

Check whether age labels conform to expectations:

```{r check}
lab <- age_labels_five(
  lower_first = 0,
  lower_last = 85,
  open_right = TRUE
)
age_diagnose(
  lab,
  no_overlap = TRUE,
  no_gap = TRUE,
  has_zero = TRUE,
  has_open_right = TRUE
)
```

`age_assert()` throws an error if a check fails.

## Create mappings

`age_mapping()` builds a mapping between sets of labels.

```{r mapping-contains}
labels_x <- c("10--14", "0--9")
labels_y <- c("0-4", "5-9", "10-14")
age_mapping(
  labels_x,
  labels_y,
  relation = "contains",
  format = "matrix"
)
```


## Period and cohort

Age group functions have period and cohort equivalents. For instance:

| Age                 | Period                 | Cohort                 |
|---------------------|------------------------|------------------------|
| `age_lower()`       | `period_lower()`       | `cohort_lower()`       |
| `age_standard()`    | `period_standard()`    | `cohort_standard()`    |
| `age_labels_five()` | `period_labels_five()` | `cohort_labels_five()` |
| `age_coarsen()`     | `period_coarsen()`     | `cohort_coarsen()`     |


## Function tasks

`agetime` functions fall into five groups, defined by their tasks:

| Task | Examples | Return value |
|-------------------------------------------|---------------------------------------------------------------|----------------------------------|
| Extract limits or identify special labels | `age_lower()`, `age_upper()`, `age_mid()`, `age_width()`, `age_is_total()` | Numeric or logical |
| Create a new series, or reformat existing labels | `age_labels_five()`, `age_standard()`, `age_extend()` | Character, or same type as `labels` |
| Recode values and set factor levels | `age_coarsen()`, `age_fill()`, `age_set_open_right()`, `age_set_order()` | Factor |
| Validate labels | `age_diagnose()`, `age_assert()` | Diagnostics, or `labels` |
| Relate two sets of labels | `age_mapping()` | Tibble or matrix |


## More information

- `?agetime` for a full list of functions
- `vignette("cookbook")` for suggestions for performing common label-related tasks
- `vignette("api-principles")` for an AI-written summary of principles underlying the interface for the package
- Package website: <https://bayesiandemography.github.io/agetime/>
