---
title: "Quick start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Demographic tables often store age, period, or cohort as **character labels** such as `"0-14"`, `"2020-2025"`, or `"100+"`.
The **agetime** package parses these labels into the numeric intervals they represent, so you can filter, recode, and validate them reliably.

This vignette introduces the main ideas using **age** labels.
Equivalent functions exist for **period** and **cohort** (`period_*()`, `cohort_*()`).

## Read intervals from labels

Given a vector of age group labels, extract lower limits, upper limits, and widths:

```{r extract}
library(agetime)

x <- c("0-4", "5-9", "100+")
age_lower(x)
age_upper(x)
age_width(x)
```

Open-ended groups (`"100+"`) and closed groups are handled consistently.
Use the lower limit to **filter** data:

```{r filter}
library(dplyr, warn.conflicts = FALSE)

df <- tibble(
  age   = c("0-14", "15-64", "65+"),
  count = c(100, 200, 50)
)

df |>
  filter(age_lower(age) >= 15)
```

## Standardize messy labels

Real data often uses inconsistent formatting.
`age_standard()` converts labels to a standard form:

```{r standard}
x <- c("5to9", "10--14", "100plus")
age_standard(x)
```

If `x` is a factor, `age_standard()` returns a factor and updates `levels()` as well as values.

## Create a regular series of labels

To generate labels from scratch, use `age_labels_*()` functions.
For example, five-year age groups from 0 to 20:

```{r labels}
age_labels_five(lower_first = 0, lower_last = 20)
```

See `?age_labels` for one-year, ten-year, and life-table variants.

## Modify to new groupings

To **widen** or **realign** age groups while ensuring every old group fits inside a new one, use `age_modify()` or the convenience functions `age_modify_five()`, `age_modify_ten()`, and `age_modify_life()`:

```{r modify}
x <- c("0-4", "5-9", "10-14", "50-54")
age_modify(x, breaks = c(0, 15, 55))
age_modify_five(x)
```

Related helpers work on factor **levels** rather than recoding values: `age_levels_fill()` adds missing age groups to the levels (see `?age_levels_fill` for one-year, five-year, ten-year, and life-table variants), and `age_levels_sort()` sorts levels by their lower limits.

## Check and assert

Before analysis, check whether a set of labels has gaps, overlaps, or other properties:

```{r check}
lab <- age_labels_five(lower_first = 0, lower_last = 85, open = TRUE)
age_check(lab,
          no_overlap = TRUE,
          no_gap = TRUE,
          include_zero = TRUE,
          include_open = TRUE)
```

Use `age_assert()` when you want a failed check to throw an error.
For individual labels, `age_is_open()` and `age_is_total()` return logical vectors indicating open-ended groups and totals.

## Map between label sets

When comparing or aggregating tables that use different age groupings, `age_mapping()` builds a mapping between two sets of labels.
By default it finds pairs with **equal** intervals; the `relation` argument can instead match groups where one **contains** the other, is **contained** in the other, or **overlaps** (see `?age_mapping` for definitions).
If `y` is omitted, `x` is mapped onto itself.

```{r mapping}
x <- c("0-4", "5-9", "10-14")
y <- c("0-9", "10-19")
age_mapping(x, y)
age_mapping(x, y, relation = "contained")
```

## Period and cohort

The same workflow applies to periods and cohorts.
Function names follow the pattern `period_*()` and `cohort_*()`:

| Task              | Age              | Period              | Cohort              |
|-------------------|------------------|---------------------|---------------------|
| Lower limit       | `age_lower()`    | `period_lower()`    | `cohort_lower()`    |
| Standardize       | `age_standard()` | `period_standard()` | `cohort_standard()` |
| Five-year labels  | `age_labels_five()` | `period_labels_five()` | `cohort_labels_five()` |
| Modify            | `age_modify()`   | `period_modify()`   | `cohort_modify()`   |

Period and cohort functions accept `x_one` and `x_multi` arguments to control how one-year and multi-year labels are interpreted; see `?period_lower` and `?cohort_lower`.

## Learn more

- `?agetime` for a full list of functions grouped by task
- Package website: <https://bayesiandemography.github.io/agetime/>
