---
title: "Quick start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Datasets often use character labels such as `"0-14"`, `"2020-2025"`, or `"1980-1990"` for age groups, periods, and cohorts.
**agetime** infers the numeric intervals underlying the labels, and provides tools for filtering, recoding, and validating these intervals.

This vignette focuses on age group labels. **agetime** functions for periods and cohorts work like the functions for ages.

## Read intervals from labels

Given a vector of age group labels, extract lower limits, upper limits, midpoints, and widths:

```{r extract}
library(agetime)

x <- c("0", "1-4", "5-9", "10+")
age_lower(x)
age_upper(x)
age_mid(x)
age_width(x)
```

Use the lower limit to filter rows:

```{r filter}
library(dplyr, warn.conflicts = FALSE)

df <- tibble(
  age   = c("0-14", "15-64", "65+"),
  count = c(100, 200, 50)
)

df |>
  filter(age_lower(age) >= 15)
```

`age_mid()` is useful for plotting.

## Find special labels

**agetime** understands open intervals and totals.

```{r filter-total-open}
df <- tibble(
  age   = c("15-64", "0-14", "65+", "All"),
  count = c(5, 2, 6, 13)
)

df |>
  filter(!age_is_total(age))

df |>
  filter(age_is_open(age))
```

## Standardize messy labels

There are many ways of formatting age group labels. `age_standard()` converts labels to a standard format:

```{r standard}
x <- c("5to9", "10--14", "100plus")
age_standard(x)
```

If `x` is a factor, `age_standard()` returns a factor where the levels are also standardized.

## Create a regular series of labels

To generate labels from scratch, use the `age_labels` functions.

```{r labels}
age_labels_five(lower_first = 0, lower_last = 20)
age_labels_life()
```

## Modify to new groupings

To widen or realign age groups, use an `age_modify` function.

```{r modify}
x <- c("0-4", "5-9", "10-14", "50-54")
age_modify(x, breaks = c(0, 15, 55))
age_modify_five(x)
```

`age_levels_fill()` adds missing age groups to factor levels, and `age_levels_sort()` sorts levels.

## Check and assert

Check whether age labels conform to expectations:

```{r check}
lab <- age_labels_five(lower_first = 0, lower_last = 85, open = TRUE)
age_check(lab,
          no_overlap = TRUE,
          no_gap = TRUE,
          include_zero = TRUE,
          include_open = TRUE)
```

With `age_assert()`, failed checks throw an error.

## Create mappings

`age_mapping()` builds a mapping between sets of labels.

By default, `age_mapping()` looks for intervals that are identical.

```{r mapping}
x <- c("10--14", "0--9")
y <- c("0-4", "5-9", "10-14")
age_mapping(x, y, return_val = "matrix")
```

But it can look for other types of relationships.

```{r mapping-contains}
x <- c("10--14", "0--9")
y <- c("0-4", "5-9", "10-14")
age_mapping(x, y, relation = "contains", return_val = "matrix")
```


## Period and cohort

Age group functions have period and cohort equivalents. For instance:

| Age                 | Period                 | Cohort                 |
|---------------------|------------------------|------------------------|
| `age_lower()`       | `period_lower()`       | `cohort_lower()`       |
| `age_standard()`    | `period_standard()`    | `cohort_standard()`    |
| `age_labels_five()` | `period_labels_five()` | `cohort_labels_five()` |
| `age_modify()`      | `period_modify()`      | `cohort_modify()`      |


## Learn more

- `?agetime` for a full list of functions grouped by task
- Package website: <https://bayesiandemography.github.io/agetime/>
