---
title: "API principles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{API principles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`agetime` functions are designed around label vectors. A label such as
`"5-9"`, `"2020-2025"`, or `"<1990"` is a character value, but it also
implies an interval. The API tries to make this distinction explicit:
some functions read intervals from existing labels, some functions create
new labels, and some functions work with factor levels.

## Domain prefixes

Most functions start with a domain:

- `age` for age groups
- `period` for time periods
- `cohort` for birth cohorts

The three domains usually share the same operation names. For example,
`age_lower()`, `period_lower()`, and `cohort_lower()` all extract lower
limits from labels.

```{r domains}
library(agetime)

age_lower(c("0-4", "5-9"))
period_lower(c("2020-2025", "2025-2030"))
cohort_lower(c("<1990", "1990-1995"))
```

A few operations are domain-specific:

- Open intervals on the left (`is_open_left`, `set_open_left`) exist for
  periods and cohorts only. Ages open on the right only.
- Life-table helpers (`labels_life`, `coarsen_life`, `fill_life`, and the
  `valid_life` diagnose check) exist for ages only.

See `?agetime` for the full inventory, including blank cells where a domain
has no equivalent.

## Operation families

Function names use the domain prefix plus an operation family.

| Family | Purpose | Return type |
|---|---|---|
| `lower`, `upper`, `width`, `mid` | Extract interval properties | Numeric vector |
| `is_open_left`, `is_open_right`, `is_total`, `is_missing`, `is_subtotal` | Identify special labels | Logical vector |
| `diagnose`, `assert` | Validate labels (factor levels by default) | List with `ok`, or `labels` / error |
| `diagnose_values`, `assert_values` | Validate observed values only | List with `ok`, or `labels` / error |
| `standard` | Convert labels to standard format | Same type and length as `labels` |
| `coarsen`, `coarsen_to` | Regroup labels into new intervals | Factor |
| `extend` | Continue a label series | Character vector or factor |
| `fill`, `set_open_left`, `set_open_right`, `set_order` | Work with factor levels | Factor |
| `labels` | Create new labels | Character vector |
| `mapping` | Map between interval sets | Tibble or matrix |

The most important return-type distinction is between labels and levels:

- The `labels` functions create character vectors of labels.
- The `coarsen`, `fill`, `set_open_left`, `set_open_right`, and `set_order`
  functions return factors, because their purpose is to recode values and
  shape factor levels.

```{r labels-vs-levels}
age_labels_five(lower_last = 20)

age_coarsen(c("10-14", "0-4"), breaks = c(0, 10, 20))
age_fill_five(c("10-14", "0-4"))
```

## Input parsing and output rendering

The main input argument is called `labels` when it is a vector of age, period,
or cohort labels. Arguments beginning with `interpret_` describe how existing
input labels should be parsed. They do not describe how newly generated output
labels should be rendered.

Age labels follow a strong demographic convention: one-year labels use the
lower boundary, and multi-year labels exclude the upper boundary. Age
functions therefore do not expose `interpret_single` or `interpret_multi`
arguments. Period and cohort labels do not have the same universal convention,
so period and cohort functions that *consume* labels (for example
`period_standard()` or `cohort_coarsen()`) let you specify how inputs should
be parsed. The period and cohort `labels` constructors do not take
`interpret_` arguments; they create new labels instead.

For example, `period_standard()` uses `interpret_multi` to interpret the input:

```{r input-parsing}
period_standard("2020-2025", interpret_multi = "exclude")
```

Label-consuming functions also take `interpret_fail`, which controls what
happens when a label cannot be parsed (`"error"`, `"warn"`, or `"silent"`).

Arguments beginning with `format_` describe how new output labels should be
rendered. Only the period and cohort `labels` constructors expose
`format_single` and `format_multi`. Age `labels` functions hard-code the
usual demographic conventions. For example, `period_labels_ten()` uses
`format_multi` to decide whether the displayed upper limit is included or
excluded:

```{r output-rendering}
period_labels_ten(
  lower_first = 2000,
  lower_last = 2020,
  format_multi = "exclude"
)
```

This convention keeps input interpretation separate from output formatting.
(The `mapping` functions use a different argument named `format`, which
chooses between a tibble and a matrix return value.)

## Boundary names

Boundary arguments name the boundary they represent.

- `breaks` are boundaries between intervals.
- `lower_first` is the lower bound of the first generated interval.
- `lower_last` is the lower bound of the last generated interval.
- `at` is the breakpoint at which intervals become open, e.g. `70` in `70+`
  or `1990` in `<1990`. The function name specifies whether intervals open on
  the left or right.

```{r boundaries}
age_labels_five(lower_first = 0, lower_last = 20, open_right = FALSE)

age_set_open_right(c("0-4", "60-64"), at = 70)

cohort_set_open_left(
  c("2000-2004", "2010-2014"),
  at = 1990,
  interpret_multi = "exclude"
)
```

The boolean arguments `open_left` and `open_right` control whether to include
open intervals at the start or end of a generated label sequence.
Age groups support `open_right` only; periods and cohorts support both
directions.

In the `labels` functions, these arguments default to explicit boolean
values because you are constructing a new label grid. In the `coarsen`
functions, the defaults are `NULL`, which preserves the open/closed topology
of `labels`: open ends are kept when present, and omitted when not. Use
`open_left = TRUE` or `open_right = TRUE` to add a structural open level
even when no value in `labels` currently belongs to that group.

## Values and levels

Functions such as `age_coarsen()`, `age_fill()`, `age_set_open_right()`, and
`age_set_order()` always return a factor. They preserve observed values where
possible, and change or add levels. By default, `age_coarsen()` infers
whether to include an open top from `labels`. Use `open_right = TRUE` to add
an open level even when no value in `labels` currently belongs to that open
group, as `age_set_open_right()` does:

```{r level-contract}
labels <- c("0-4", "60-64")
age_coarsen(labels, breaks = c(0, 50, 70), open_right = TRUE)
```

The `diagnose` and `assert` functions follow the same levels contract: for a
factor they validate levels, including unused ones. Use the
`diagnose_values` and `assert_values` functions when you want to check only
observed values (for example in a grouped pipeline). The `assert` functions
return `labels` invisibly when checks pass, so they can sit in a pipeline.

## More information

- See `vignette("agetime")` for a quick start.
- See `vignette("cookbook")` for task-oriented recipes.
- See `?agetime` for a complete function list grouped by task.
