---
title: "API principles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{API principles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**agetime** functions are designed around label vectors. A label such as
`"5-9"`, `"2020-2025"`, or `"<1990"` is a character value, but it also
implies an interval. The API tries to make this distinction explicit:
some functions read intervals from existing labels, some functions create
new labels, and some functions work with factor levels.

## Domain prefixes

Most functions start with a domain:

- `age_*()` for age groups.
- `period_*()` for time periods.
- `cohort_*()` for birth cohorts.

The three domains usually share the same operation names. For example,
`age_lower()`, `period_lower()`, and `cohort_lower()` all extract lower
limits from labels.

```{r domains}
library(agetime)

age_lower(c("0-4", "5-9"))
period_lower(c("2020-2025", "2025-2030"))
cohort_lower(c("<1990", "1990-1995"))
```

## Operation families

Function names use the domain prefix plus an operation family.

| Family | Purpose | Return type |
|---|---|---|
| `*_lower()`, `*_upper()`, `*_width()`, `*_mid()` | Extract interval properties | Numeric vector |
| `*_is_open()`, `*_is_total()` | Identify special labels | Logical vector |
| `*_check()`, `*_assert()` | Validate labels | Diagnostics or error |
| `*_standard()` | Convert labels to standard format | Same shape as `labels` |
| `*_modify()` | Regroup labels into new intervals | Same shape as `labels` |
| `*_extend()` | Continue a label series | Character vector or factor |
| `*_levels_*()` | Work with factor levels | Factor |
| `*_labels_*()` | Create new labels | Character vector |
| `*_mapping()` | Map between interval sets | Matrix or tibble |

The most important return-type distinction is between labels and levels:

- `*_labels_*()` functions create character vectors of labels.
- `*_levels_*()` functions return factors, because their purpose is to
  create or reorder factor levels.

```{r labels-vs-levels}
age_labels_five(lower_last = 20)

age_levels_fill_five(c("10-14", "0-4"))
```

## Input parsing and output rendering

The main input argument is called `labels` when it is a vector of age, period,
or cohort labels. Arguments beginning with `interpret_` describe how existing
input labels should be parsed. They do not describe how newly generated output
labels should be rendered.

Age labels follow a strong demographic convention: one-year labels use the
lower boundary, and multi-year labels exclude the upper boundary. Age
functions therefore do not expose `interpret_single` or `interpret_range`
arguments. Period and cohort labels do not have the same universal convention,
so period and cohort functions let you specify how input labels should be
parsed.

For example, `period_standard()` uses `interpret_range` to interpret the input:

```{r input-parsing}
period_standard("2020-2025", interpret_range = "exclude")
```

Arguments beginning with `format_` describe how new output labels should be
rendered. For example, `period_labels_ten()` uses `format_range` to decide
whether the displayed upper limit is included or excluded:

```{r output-rendering}
period_labels_ten(
  lower_first = 2000,
  lower_last = 2020,
  format_range = "exclude"
)
```

This convention is intended to keep input interpretation separate from
output formatting.

## Boundary names

Boundary arguments name the boundary they represent.

- `breaks` are boundaries between intervals.
- `lower_first` is the lower bound of the first generated interval.
- `lower_last` is the lower bound of the last generated interval.
- `lower_open` is the lower bound of a right-open age group such as `70+`.
- `upper_open` is the upper bound of a left-open cohort such as `<1990`.

```{r boundaries}
age_labels_five(lower_first = 0, lower_last = 20, open = FALSE)

age_levels_set_open(c("0-4", "60-64"), lower_open = 70)

cohort_levels_set_open(
  c("2000-2004", "2010-2014"),
  upper_open = 1990,
  interpret_range = "exclude"
)
```

The boolean argument `open` is used where the question is whether to include
an open interval at all. Boundary names such as `lower_open` and
`upper_open` are used where the question is where the open interval starts
or ends.

## Values and levels

Functions that modify values, such as `age_modify()`, return values with the
same length as `labels`. If `labels` is a factor, they also update levels.

Functions in the `*_levels_*()` family always return a factor. They preserve
observed values where possible, and change or add levels. This is why
`age_levels_set_open()` can add a `70+` level even when no value in `labels`
currently belongs to that open group.

```{r level-contract}
labels <- c("0-4", "60-64")
age_levels_set_open(labels, lower_open = 70)
```

## More information

- See `vignette("agetime")` for a quick start.
- See `?agetime` for a complete function list grouped by task.
