---
title: "Cookbook"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Cookbook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows how to use functions from the  `agetime` package to perform tasks that often arise when dealing with age, period, and cohort labels.


```{r libs}
library(agetime)
library(dplyr, warn.conflicts = FALSE)
```

# Preparing data


## Drop totals, missing, or isolate open intervals

### Task

Remove total or missing categories, or keep only open-ended intervals,
before summing or plotting.

### Solution

Identify special labels with `*_is_total()`, `*_is_missing()`, and
`*_is_open_*()`, then filter.

```{r filter-special}
df <- tibble(
  age = c("0-14", "15-64", "65+", "All", "not stated"),
  period = c("2015-2020", "2020-2025", "<2015", "Total", NA),
  count = c(10, 20, 5, 35, 1)
)

df |>
  filter(
    !age_is_total(age),
    !age_is_missing(age),
    !age_is_open_right(age)
  )

df |>
  filter(period_is_open_left(period))
```

### Discussion

Totals, missing values, and open ends are often stored as ordinary
character labels. Treat them explicitly so they are not mixed into
ordinary age or period sums.

### See Also

- `cohort_is_open_left()`, `cohort_is_open_right()`
- `period_is_missing()`, `cohort_is_missing()`
- [Impose an open end on factor levels](#impose-an-open-end-on-factor-levels)

## Standardize messy labels before analysis

### Task

Convert inconsistently formatted labels to a single default style.

### Solution

Use `age_standard()` (or the period/cohort equivalent) inside
`dplyr::mutate()`.

```{r standard}
df <- tibble(
  age = c("5to9", "10--14", "100plus", "all"),
  count = c(11, 9, 2, 22)
)

df |>
  mutate(age = age_standard(age))
```

### Discussion

If `age` is a factor, levels are standardized as well as values.

### See Also

- `vignette("api-principles")` for the default formats
- [Validate incoming labels](#validate-incoming-labels)



## Continue a period or cohort series

### Task

Extend an existing sequence of periods (or cohorts) forward in time.

### Solution

Use `period_extend()` (or `cohort_extend()`).

```{r extend}
recent <- c("2020-2025", "2025-2030")
period_extend(recent, n = 2)
```

### See Also

- `period_sort()`, `cohort_sort()` before extending a factor whose
  levels are out of order
- [Build a complete age or period skeleton](#build-a-complete-age-or-period-skeleton)

## Impose an open end on factor levels

### Task

Add an open-ended level (such as `70+` or `<2000`) for publication or
modelling, even when no observation falls in that group yet.

### Solution

Use `age_set_open_right()` or `cohort_set_open_left()` (and the other
domain-specific open helpers).

```{r set-open}
age_set_open_right(c("0-4", "60-64"), at = 70)

cohort_set_open_left(c("2000-2004", "2010-2014"), at = 2000)
```

### See Also

- [Drop totals, missing, or isolate open intervals](#drop-totals-missing-or-isolate-open-intervals)
- `period_set_open_left()`, `period_set_open_right()`

## Order levels for tables and plots

### Task

Put factor levels in chronological or age order.

### Solution

```{r sort}
messy <- factor(
  c("10-14", "0-4", "5-9"),
  levels = c("10-14", "0-4", "5-9")
)
age_sort(messy)
```

### Discussion

Combine with fill or set-open when you need a complete, ordered set of
levels.

### See Also

- [Build a complete age or period skeleton](#build-a-complete-age-or-period-skeleton)
- [Impose an open end on factor levels](#impose-an-open-end-on-factor-levels)



## Aggregate to wider groups

### Task

Collapse fine age groups to a coarser classification and sum counts.

### Solution

Coarsen labels, then summarise with dplyr.

```{r coarsen-aggregate}
deaths <- tibble(
  age = c("0", "1", "2", "5", "6", "10", "11"),
  n = c(3, 1, 2, 4, 5, 6, 7)
)

deaths |>
  mutate(age = age_coarsen_five(age)) |>
  group_by(age) |>
  summarise(n = sum(n), .groups = "drop")
```

### Discussion

For irregular breaks, pass `breaks` to `age_coarsen()`. Life-table
groupings are available via `age_coarsen_life()`.

```{r coarsen-breaks}
tibble(
  age = c("0-4", "5-9", "10-14", "15-19"),
  n = c(10, 20, 15, 12)
) |>
  mutate(age = age_coarsen(age, breaks = c(0, 15, 20))) |>
  group_by(age) |>
  summarise(n = sum(n), .groups = "drop")
```

### See Also

- [Map between two classifications](#map-between-two-classifications)
- `age_coarsen_ten()`, `period_coarsen_five()`

## Build a complete age or period skeleton

### Task

Ensure every planned age group appears in a table, even when some have
no observations.

### Solution

Generate the full set of labels with `*_labels_*()`, then join observed
data onto that skeleton.

```{r skeleton}
skeleton <- tibble(
  age = age_labels_five(lower_first = 0, lower_last = 20, open_right = FALSE)
)

observed <- tibble(
  age = c("0-4", "10-14"),
  n = c(30, 25)
)

skeleton |>
  left_join(observed, by = "age")
```

### Discussion

When the problem is missing **factor levels** (for tables or models)
rather than missing rows, use `age_fill()` or `age_fill_five()`:

```{r fill-levels}
x <- factor(c("0-4", "10-14"), levels = c("0-4", "10-14"))
age_fill_five(x)
```

### See Also

- [Order levels for tables and plots](#order-levels-for-tables-and-plots)
- `age_labels_life()`, `period_labels_five()`


# Analysing data

## Filter rows by age, period, or cohort

### Task

Select rows whose age groups fall in a given range.

### Solution

Use lower or upper limits with `dplyr::filter()`.

```{r filter-limits}
df <- tibble(
  age = c("0-14", "15-64", "65+", "Total"),
  count = c(100, 200, 50, 350)
)

df |>
  filter(age_lower(age) >= 15, age_upper(age) <= 65)
```

### See Also

- [Drop totals, missing, or isolate open intervals](#drop-totals-missing-or-isolate-open-intervals)
- `age_mid()`, `age_width()`


## Plot against interval midpoints

### Task

Place age groups on a numeric axis for a plot.

### Solution

Use `age_mid()` (including honorary midpoints for open intervals).

```{r mid-plot, fig.width = 5, fig.height = 3.5}
rates <- tibble(
  age = c("0-4", "5-9", "10-14", "15-19", "20+"),
  rate = c(0.02, 0.01, 0.015, 0.03, 0.05)
)

plot(
  age_mid(rates$age),
  rates$rate,
  type = "b",
  xlab = "Age (midpoint)",
  ylab = "Rate"
)
```

### See Also

- [Measure interval length](#measure-interval-length)
- `period_mid()`, `cohort_mid()`



## Measure interval length

### Task

Turn interval labels into widths for exposure-style calculations.

### Solution

```{r width}
pop <- tibble(
  age = c("0-4", "5-9", "10-14", "85+"),
  person_years = c(500, 480, 470, 200)
)

pop |>
  mutate(
    width = age_width(age),
    approx_midyear_pop = person_years / width
  )
```

### Discussion

Open intervals have undefined width (`NA`).

### See Also

- [Plot against interval midpoints](#plot-against-interval-midpoints)
- `period_width()`, `cohort_width()`








# Programming

## Validate incoming labels

### Task

Check that a label vector meets expectations, either for a report or to
stop a pipeline.

### Solution

Use `*_check()` for diagnostics and `*_assert()` to fail fast.

```{r validate}
lab <- c("2020-2025", "2025-2030", "2030-2035")
period_check(lab, no_gap = TRUE, no_overlap = TRUE)

pop <- tibble(
  age = c("0-4", "5-9", "10-14"),
  count = c(100, 110, 105)
)

pop |>
  mutate(age = age_assert(age, has_zero = TRUE, no_gap = TRUE))
```

### Discussion

`*_check()` returns a list with an `ok` flag and details. `*_assert()`
returns its input invisibly when checks pass, so it fits in a mutate
pipeline.

### See Also

- `?age_check` for the full set of conditions



## Map between two classifications

### Task

Relate a coarse period classification to a finer one and reallocate
counts.

### Solution

Build a mapping with `period_mapping()`, then join and summarise.

```{r mapping}
coarse <- c("2020-2030", "2030-2040")
fine <- c("2020-2025", "2025-2030", "2030-2035", "2035-2040")

map <- period_mapping(
  labels_x = coarse,
  labels_y = fine,
  relation = "contains"
)
map

fine_counts <- tibble(
  period = fine,
  n = c(10, 12, 9, 11)
)

map |>
  left_join(fine_counts, by = c("y" = "period")) |>
  group_by(x) |>
  summarise(n = sum(n), .groups = "drop")
```

### Discussion

Other relations include `"equals"`, `"is-contained-in"`, and
`"overlaps-with"`. The matrix form of `format = "matrix"` is useful when
you need a dense indicator grid.

### See Also

- [Aggregate to wider groups](#aggregate-to-wider-groups)
- `age_mapping()`, `cohort_mapping()`
