---
title: "Cookbook"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Cookbook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette shows how to use functions from package `agetime` to carry out common tasks when working with age, period, and cohort data. 

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r libs}
library(agetime)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
set.seed(0)
```


# Create and modify labels

## Create labels from scratch

### Problem

Create a new set of age group, cohort, or period labels.

### Solution

For labels with regular widths, use function such as `age_labels_ten()` or `period_labels_five()`.

```{r labels-from-scratch-regular}
age_labels_ten()
period_labels_five(
  lower_first = 2001,
  lower_last = 2031
)
```

For labels with irregular widths, use `age_labels()`, `period_labels()`, or `cohort_labels()`.

```{r labels-from-scratch-irregular}
age_labels(
  breaks = c(0, 20, 65, 100),
)
cohort_labels(
  breaks = c(2000, 2010, 2030),
  open_left = TRUE
)
```

### Discussion

- The help for `age_labels()`, `period_labels()`, and `cohort_labels()` describes official `agetime` style.

### See Also

TODO - COMPLETE

## Convert labels to a default format {#standardize}

### Problem

Convert age, period, or cohort labels to the official `agetime` style.

### Solution

Use `age_standard()`, `period_standard()`, or `cohort_standard()`.

```{r standard}
df <- tibble(
  age = c("5to9", "10--14", "100plus", "infants"),
  count = c(11, 9, 2, 22)
)

df |>
  mutate(age = age_standard(age))
```

### Discussion

If the `labels` argument to `age_standard()`, `period_standard()`, or `cohort_standard()` is a factor, then the factor levels are standardized along with the values.


### See Also

- The help for `age_standard()`, `period_standard()`, and `cohort_standard()` describe official `agetime` style.


## Create period labels for a forecast

### Problem

Take a sequence of periods and extend it forward in time.

### Solution

Use `period_extend()`.

```{r extend}
recent <- c("2020-2025", "2025-2030")
period_extend(recent, n = 2)
```

### See Also

- `age_extend()` and `cohort_extend()` do the same thing for age groups and cohorts.

## Specify an open age group, period, or cohort

### Problem

Move the starting point of existing open age groups, periods, or cohorts, or specify a new open age group, period, or cohort.

An open age group, period, or cohort is one that has no lower limit (e.g. `"<2000"`) or no upper limit (e.g. `"85+"`).

### Solution

Use 
 
 - `age_set_open_right()`
 - `period_set_open_left()`
 - `period_set_open_right()`
 - `cohort_set_open_left()`
 - `cohort_set_open_right()`.

```{r set-open}
labels_age <- c("0-4", "60-64", "90+")
age_set_open_right(labels_age, at = 80)

labels_cohort <- c("2000-2004", "2010-2014")
labels_cohort |>
  cohort_set_open_left(at = 1990) |>
  table()
```

### Discussion

The `*_set_open_*()` functions change the levels of `labels` as well as the values. If `labels` is not a factor, the `*_set_open_*()` functions turn it into one.

### See Also

TODO - COMPLETE

## Create an age variable that sorts properly {#sort-age-permanently}

### Problem

Raw age group labels don't sort properly:

```{r sort-age-wrong}
df <- tibble(
  age = c("10-14", "0-4", "5-9"),
  count = c(33, 7, 15)
)

df |>
  arrange(age) ## "10-14" comes before "5-9"!!
```

### Solution

Function `age_set_order()` creates a factor with levels in the right order.

```{r set-order}
labels_unordered <- c("10-14", "0-4", "5-9")
age_set_order(labels_unordered)
```

### Discussion

If `labels` is not a factor, `age_set_order()` turns it into one.

### See Also

- `period_set_order()` and `cohort_set_order()` permanently modify period and cohort variables.
- [Order by age when working interactively](#sort-age-interactive)


## Build a complete age or period skeleton

### Problem

Ensure every planned age group appears in a table, even when some have
no observations.

### Solution

Generate the full set of labels with `*_labels_*()`, then join observed
data onto that skeleton.

```{r skeleton}
skeleton <- tibble(
  age = age_labels_five(lower_first = 0, lower_last = 20, open_right = FALSE)
)

observed <- tibble(
  age = c("0-4", "10-14"),
  n = c(30, 25)
)

skeleton |>
  left_join(observed, by = "age")
```

### Discussion

When the problem is missing **factor levels** (for tables or models)
rather than missing rows, use `age_fill()` or `age_fill_five()`:

```{r fill-levels}
x <- factor(c("0-4", "10-14"), levels = c("0-4", "10-14"))
age_fill_five(x)
```

### See Also

- [Order levels for tables and plots](#order-levels-for-tables-and-plots)
- `age_labels_life()`, `period_labels_five()`


## Validate incoming labels

### Problem

Check that a label vector meets expectations, either for a report or to
stop a pipeline.

### Solution

Use `*_check()` for diagnostics and `*_assert()` to fail fast.

```{r validate}
lab <- c("2020-2025", "2025-2030", "2030-2035")
period_check(lab, no_gap = TRUE, no_overlap = TRUE)

pop <- tibble(
  age = c("0-4", "5-9", "10-14"),
  count = c(100, 110, 105)
)

pop |>
  mutate(age = age_assert(age, has_zero = TRUE, no_gap = TRUE))
```

### Discussion

`*_check()` returns a list with an `ok` flag and details. `*_assert()`
returns its input invisibly when checks pass, so it fits in a mutate
pipeline.

### See Also

- `?age_check` for the full set of conditions




# Analyse data

## Filter rows by age, period, or cohort

### Problem

Select rows based on ages, periods, or cohorts.

### Solution

Filter on lower or upper limits.

```{r filter-limits}
df <- tibble(
  age = c("0-14", "15-64", "65+", "Total"),
  period = c("2017", "2033", "2050+", "2020-2030"),
  count = c(100, 200, 50, 350)
)

df |>
  filter(age_lower(age) >= 15)

df |>
  filter(period_upper(period) <= 2030)
```

### See Also

TODO - ADD LINKS


## Drop totals from a dataset {#drop-totals}

### Problem

Remove rows containing age, period, or cohort totals.

### Solution

Use `age_is_total()`, `period_is_total()`, or `cohort_is_total()` to identify rows with totals and then filter. 

```{r drop-totals}
df <- tibble(
  age = c("0-14", "15-64", "65+", "All"),
  count = c(10, 20, 5, 35)
)

df |>
  filter(!age_is_total(age))
```

### Discussion

`age_is_total()`, `period_is_total()`, and `cohort_is_total()` recognize synonyms for totals (e.g. `"all"`) and ignore case (e.g. `"tOTaL"`).

### See Also

- [Drop NA and not-stated from a dataset](#drop-missing)


## Drop "Not stated" categories from a dataset {#drop-missing}

### Problem

Remove rows where data on age, period, or cohort is not stated or missing.

### Solution

Use `age_is_missing()`, `period_is_missing()`, or `cohort_is_missing()` to identify rows with not stated or missing and then filter. 

```{r drop-missing}
df <- tibble(
  time = c("2005", "unknown", "2005-2010", NA),
  count = c(100, 2, 55, 12)
)

df |>
  filter(!period_is_missing(time))
```

### Discussion

`age_is_missing()`, `period_is_missing()`, and `cohort_is_total()` look for `NA`s, but also for text equivalents, such as `"Not stated"`, `"Don't know"` or `"n\a"`.

### See Also

- [Drop totals from a dataset](#drop-totals)


## Order by age when working interactively {#sort-age-interactive}

### Problem

Sort rows by age group as a one-off operation.

Note that sorting on raw age group labels can give the wrong result

```{r sort-age-wrong-interactive}
df <- tibble(
  age_group = c("5-9", "0-4", "10-14"),
  count = c(33, 7, 15)
)

df |>
  arrange(age_group) ## "10-14" comes before "5-9"!!
```

### Solution

Sort on lower or upper limits.

```{r sort-age-lower}
df <- tibble(
  age_group = c("5-9", "0-4", "10-14"),
  count = c(33, 7, 15)
)

df |>
  arrange(age_lower(age_group))
```

### See Also

- [Create an age variable that sorts properly](#sort-age-permanently)


## Include levels that don't appear in the data

### Problem

Some age groups, periods, or cohorts don't appear in the data. Tabulations and other calculations don't account for these levels, leading to gaps in the output.

```{r fill-gaps-problem}
df <- tibble(
  age = c("0-4", "10-14", "0-4"),
  respondents = c(10, 20, 5)
)

## misses "5-9"
df |>
  count(age, wt = respondents)
```

### Solution

Use `age_fill()`, `period_fill()`, or `cohort_fill()`.

```{r fill-gaps-solution}
df <- tibble(
  age = c("0-4", "10-14", "0-4"),
  respondents = c(10, 20, 5)
)

## includes 5-9
df |>
  mutate(age = age_fill(age)) |>
  count(age, wt = respondents, .drop = FALSE) 
```

### Discussion

`age_fill()`, `period_fill()`, and `cohort_fill()` change the levels of `labels`, not the values. If `labels` is not a factor, `age_fill()`, `period_fill()`, and `cohort_fill()` turn it into one.

Tidyverse functions such as `count()` and `summarise()` drop unused levels by default. To include these levels, set `.drop` to `FALSE`.

### See Also

- `age_set_order()`, `period_set_order()`, `cohort_set_order()` put levels in the right order.


## Aggregate to broader age groups

### Problem

The data use one-year age groups. You want to aggregate to five-year age groups.

### Solution

Coarsen labels, then aggregate.

```{r coarsen-aggregate}
deaths <- tibble(
  age = c("0", "1", "2", "5", "6", "10", "11"),
  n = c(3, 1, 2, 4, 5, 6, 7)
)

deaths |>
  mutate(age = age_coarsen_five(age)) |>
  count(age, wt = n)
```

### Discussion

For irregular age groups, pass `breaks` to `age_coarsen()`. 
To aggregate to life-table groupings, use `age_coarsen_life()`.

### See Also

- [Recode labels into another classification](#recode-into-classification)
- `period_coarsen_five()`, `cohort_coarsen_five()`
- `period_coarsen()`, `cohort_coarsen()`


## Use age, period, or cohort on the x-axis of a plot {#plot-nicer}

### Problem

Plots that use non-numeric age groups, periods, or cohorts on the x-axis often look terrible, with overlapping labels and points instead of smooth curves. Sometimes ages are also in the wrong order, e.g. "100+" after "10-14" and "5-9" after "50-54".

```{r mid-plot-problem , fig.width = 5, fig.height = 3.5}
df <- tibble(
  age = age_labels_five(),
  rate = 10 + 0.1 * (1:21)^2 + rnorm(21)
)

ggplot(df, aes(x = age, y = rate)) +
  geom_point()
```



### Solution

Convert age, period, or cohort to a continuous variable by using `age_mid()`, `period_mid()`, or `cohort_mid()` to calculate midpoints.

```{r mid-plot-solution, fig.width = 5, fig.height = 3.5}
ggplot(df, aes(x = age_mid(age), y = rate)) +
  geom_line()
```

### See Also

- [Create an age variable that sorts properly](#sort-age-permanently)
- `period_mid()`, `cohort_mid()`



## Calculate annual rates

### Problem

Calculate annual rates from data with multi-year periods

### Solution

Use `period_width()`.

```{r width}
df <- tibble(
  period = c("2010-2019", "2020-2023", "2024-2030"),
  departures = c(2000, 50, 3000)
)

df |>
  mutate(
    n_year = period_width(period),
    annual_rate = departures / n_year
  )
```

### Discussion

The width of an open interval is undefined.

### See Also

- [Plot against interval midpoints](#plot-against-interval-midpoints)
- `age_width()`, `cohort_width()`


## Build a mapping between age groups

### Problem

Create lookup table showing the relationship between two age classifications.

### Solution

Use `age_mapping()`.

```{r mapping}
classif_fine <- age_labels_life()
classif_broad <- age_labels(breaks = c(0, 15, 65, 85))
age_mapping(
  labels_x = classif_fine,
  labels_y = classif_broad,
  relation = "is-contained-in"
)
```

### Discussion

Other possible values for `relation` are `"equals"`, `"contains"`, and
`"overlaps-with"`.

### See Also

- [Recode labels into another classification](#recode-into-classification)
- [Aggregate to broader age groups](#aggregate-to-broader-age-groups)
- `period_mapping()`, `cohort_mapping()`


## Recode labels into another classification {#recode-into-classification}

### Problem

Observed labels use a fine classification. You already have a broader
classification as a label vector, and need to recode or aggregate to those
broader labels.

### Solution

Build a fine-to-broad lookup with `age_mapping()`, then join it to the data.

```{r recode-classification}
fine <- tibble(
  age = c("0", "1", "5", "10", "11"),
  n = c(3, 1, 4, 6, 7)
)
broad <- c("0-4", "5-9", "10-14")

map <- age_mapping(
  labels_x = fine$age,
  labels_y = broad,
  relation = "is-contained-in"
)

fine |>
  left_join(map, by = c("age" = "x")) |>
  count(y, wt = n)
```

### Discussion

Prefer `age_coarsen()` or `age_coarsen_five()` when the target is a regular
break grid or fixed width, not an existing label vector.

`relation = "is-contained-in"` assumes each fine interval nests in exactly
one broad interval. If some fine labels are missing from the mapping or
appear more than once, inspect `map` before aggregating.

### See Also

- [Aggregate to broader age groups](#aggregate-to-broader-age-groups)
- [Build a mapping between age groups](#build-a-mapping-between-age-groups)
- `period_mapping()`, `cohort_mapping()`
